---
title: "Fitting Markov Modulated Point Process Models"
output: rmarkdown::html_vignette
author: Sally Sun, Owen G. Ward.
vignette: >
  %\VignetteIndexEntry{Fitting Markov Modulated Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(100) # to make it reproducible
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(ppdiag)
library(rstan)
# library(cmdstanr)
library(tidyverse)
library(bayesplot)
```


We include functions for fitting standard Hawkes and Poisson
point processes to data in `ppdiag`. However, currently, we do not
include fitting algorithms for Markov modulated point processes,
as these rely on the use of `rstan` or (more recently), `cmdstanr` to
perform Bayesian inference for the model parameters.

Here we provide a instructions on how to fit these models so 
that the can easily be used in conjunction with the diagnostic tools
of  `ppdiag`.

# Installing RStan or Cmdstanr

- Instructions for installing RStan can be found at https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started 
- Instructions for installing CmdStanR can be found at https://mc-stan.org/cmdstanr/articles/cmdstanr.html. As mentioned in this
link there are some differences between RStan and CmdStanR. In our experience,
CmdStanR does seem to be maybe more reliable and less prone to crashing.
- If you are having trouble installing or using Stan there is lots of great help
available at https://discourse.mc-stan.org/, where someone else has probably
already answered your question!


# The Stan code for MMPP/MMHP

We first include the stan code to fit each of these Markov modulated
models to a temporal point process.

```{r mmpp-stan-code}
mmpp_stan_code <- "
data{
  int<lower=1> num_events; //maximum of number of events for each pair each window => max(unlist(lapply(return_df$event.times,length))))
  vector[num_events+1] time_matrix; // include termination time in the last entry
}
parameters{
  real<lower=0> lambda0; //baseline rate for each pair
  real<lower=0> c; //baseline rate for each pair
  real<lower=0,upper=1> w1; //CTMC transition rate
  real<lower=0,upper=1> w2; //CTMC transition rate
}
transformed parameters{
  real<lower=0> q1;
  real<lower=0> q2;
  q1 = (lambda0).*w1;
  q2 = (lambda0).*w2;
}
model{
  real integ; // Placeholder variable for calculating integrals
  row_vector[2] forward[num_events]; // Forward variables from forward-backward algorithm
  row_vector[2] forward_termination; 
  row_vector[2] probs_1[num_events+1]; // Probability vector for transition to state 1 (active state)
  row_vector[2] probs_2[num_events+1]; // Probability vector for transition to state 2 (inactive state)
  vector[num_events+1] interevent;
  
  //priors
  c ~ lognormal(0,1);
  w1 ~ beta(0.5,0.5);
  w2 ~ beta(0.5,0.5);
  
  lambda0 ~ gamma(1, 1);
  interevent = time_matrix;
  // ---- prepare for forward algorithm
  // --- log probability of Markov transition logP_ij(t)
  for(n in 1:(num_events + 1)){
    probs_1[n][1] = log(q2/(q1+q2)+q1/(q1+q2)*exp(-(q1+q2)*interevent[n])); //1->1
    probs_2[n][2] = log(q1/(q1+q2)+q2/(q1+q2)*exp(-(q1+q2)*interevent[n])); //2->2
    probs_1[n][2] = log1m_exp(probs_2[n][2]); //2->1
    probs_2[n][1] = log1m_exp(probs_1[n][1]); //1->2
  }
  //consider n = 1
  integ = interevent[1]*lambda0;
  forward[1][1] = log_sum_exp(probs_1[1]) + log(lambda0*(1+c)) - integ*(1+c); 
  forward[1][2] = log_sum_exp(probs_2[1]) + log(lambda0) - integ; 
  
  if(num_events>1){
    for(n in 2:num_events){
      integ = interevent[n]*lambda0;
      forward[n][1] = log_sum_exp(forward[n-1] + probs_1[n]) + log(lambda0*(1+c))- integ*(1+c);
      forward[n][2] = log_sum_exp(forward[n-1] + probs_2[n]) + log(lambda0) - integ;
    }
  }
  
  integ = interevent[num_events]*lambda0;
  forward_termination[1] = log_sum_exp(forward[num_events] + probs_1[num_events]) - integ*(1+c);
  forward_termination[2] = log_sum_exp(forward[num_events] + probs_2[num_events]) - integ;
  
  target += log_sum_exp(forward_termination);
}
"

```

```{r mmhp-stan-code}
mmhp_stan_code <- "
data{
  int<lower=1> num_events; //number of events
  vector[num_events+1] time_matrix; // include termination time as last entry
}
parameters{
  real<lower=0> lambda0; //baseline rate for each pair
  real<lower=0> w_lambda;
  real<lower=0, upper=1> w_q1; //CTMC transition rate
  real<lower=0, upper=1> w_q2; //
  real<lower=0> alpha;
  real<lower=0> beta_delta;
  real<lower=0,upper=1> delta_1; // P(initial state = 1)
}
transformed parameters{
  real<lower=0> lambda1;
  real<lower=0> q1;
  real<lower=0> q2;
  row_vector[2] log_delta;
  real<lower=0> beta;
  lambda1 = (lambda0).*(1+w_lambda); 
  q2 = (lambda0).*w_q2;
  q1 = (lambda0).*w_q1;
  log_delta[1] = log(delta_1);
  log_delta[2] = log(1-delta_1);
  beta = alpha*(1+beta_delta);
}
model{
  real integ; // Placeholder variable for calculating integrals
  row_vector[2] forward[num_events]; // Forward variables from forward-backward algorithm
  row_vector[2] forward_termination; // Forward variables at termination time
  row_vector[2] probs_1[num_events+1]; // Probability vector for transition to state 1 (active state)
  row_vector[2] probs_2[num_events+1]; // Probability vector for transition to state 2 (inactive state)
  row_vector[2] int_1[num_events+1]; // Integration of lambda when state transit to 1 (active state)
  row_vector[2] int_2[num_events+1]; // Integration of lambda when state transit to 2 (inactive state)
  real R[num_events+1]; // record variable for Hawkes process
  vector[num_events+1] interevent;
  real K0;
  real K1;
  real K2;
  real K3;
  real K4;
  real K5;
  //priors
  w_lambda ~ gamma(1,1);
  alpha ~ gamma(1,1);//lognormal(0,1);
  beta_delta ~ lognormal(0,2);//normal(0,10);
  //delta_1 ~ beta(2,2);
  w_q1 ~ beta(2,2);
  w_q2 ~ beta(2,2);
  

  lambda0 ~ gamma(1,1);
  interevent = time_matrix;
  if(num_events==0){ // there is no event occured in this period
    //--- prepare for forward calculation
    probs_1[1][1] = log(q2/(q1+q2)+q1/(q1+q2)*exp(-(q1+q2)*interevent[1])); //1->1
    probs_2[1][2] = log(q1/(q1+q2)+q2/(q1+q2)*exp(-(q1+q2)*interevent[1])); //2->2
    probs_1[1][2] = log1m_exp(probs_2[1][2]); //2->1
    probs_2[1][1] = log1m_exp(probs_1[1][1]); //1->2
    R[1] = 0;
    K0 = exp(-(q1+q2)*interevent[1]);
    K1 = (1-exp(-(q1+q2)*interevent[1]))/(q1+q2);
    K2 = (1-exp(-(q1+q2)*interevent[1]))/(q1+q2);
    int_1[1][1] = ((q2^2*lambda1+q2*q1*lambda0)*interevent[1] +
                     (q1^2*lambda1+q2*q1*lambda0)*K0*interevent[1] +
                     (lambda1-lambda0)*q2*q1*K1 + (lambda1-lambda0)*q2*q1*K2)/(q1+q2)^2/exp(probs_1[1][1]); //1->1
    int_1[1][2] = ((q2^2*lambda1+lambda0*q1*q2)*interevent[1] -
                     (lambda1*q1*q2+lambda0*q2^2)*K0*interevent[1] +
                     (lambda0-lambda1)*q2^2*K1 + (lambda1-lambda0)*q1*q2*K2)/(q1+q2)^2/exp(probs_1[1][2]); //2->1
    int_2[1][1] = ((q1*q2*lambda1+q1^2*lambda0)*interevent[1] -
                     (q1^2*lambda1+q1*q2*lambda0)*K0*interevent[1] +
                     (lambda1-lambda0)*q1^2*K1 + q1*q2*(lambda0-lambda1)*K2)/(q1+q2)^2/exp(probs_2[1][1]); //1->2
    int_2[1][2] = ((q1*q2*lambda1+lambda0*q1^2)*interevent[1] +
                     (q1*q2*lambda1+lambda0*q2^2)*K0*interevent[1] +
                     (lambda0-lambda1)*q1*q2*K1 + (lambda0-lambda1)*q1*q2*K2)/(q1+q2)^2/exp(probs_2[1][2]); //2->2
    
    forward_termination[1] = log_sum_exp(log_delta + probs_1[1] - int_1[1]);
    forward_termination[2] = log_sum_exp(log_delta + probs_2[1] - int_2[1]);
    target += log_sum_exp(forward_termination);
    //target += -lambda0*interevent[1]*delta_1-lambda1*interevent[1]*(1-delta_1);
  }else{ // there is event occured
    // ---- prepare for forward algorithm
    // --- log probability of Markov transition logP_ij(t)
    for(n in 1:(num_events + 1)){ //changed this
      probs_1[n][1] = log(q2/(q1+q2)+q1/(q1+q2)*exp(-(q1+q2)*interevent[n])); //1->1
      probs_2[n][2] = log(q1/(q1+q2)+q2/(q1+q2)*exp(-(q1+q2)*interevent[n])); //2->2
      probs_1[n][2] = log1m_exp(probs_2[n][2]); //2->1
      probs_2[n][1] = log1m_exp(probs_1[n][1]); //1->2
    }
    
    // --- R for Hawkes
    R[1] = 0;
    for(n in 2:(num_events + 1)){ // and this
      R[n] = exp(-beta*interevent[n])*(R[n-1]+1);
    }
    
    // Integration of lambda
    for(n in 1:(num_events)){ //and this
      K0 = exp(-(q1+q2)*interevent[n]);
      K1 = (1-exp(-(q1+q2)*interevent[n]))/(q1+q2);
      K2 = (1-exp(-(q1+q2)*interevent[n]))/(q1+q2);
      K3 = R[n]*(exp(beta*interevent[n])-1)/beta;
      K4 = R[n]*(1-exp(-(beta+q1+q2)*interevent[n]))*exp(beta*interevent[n])/(beta+q1+q2);
      K5 = R[n]*(1-exp(-(q1+q2-beta)*interevent[n]))/(q1+q2-beta);
      int_1[n][1] = ((q2^2*lambda1+q2*q1*lambda0)*interevent[n] +
                       (q1^2*lambda1+q2*q1*lambda0)*K0*interevent[n] +
                       (lambda1-lambda0)*q2*q1*K1 + (lambda1-lambda0)*q2*q1*K2 +
                       alpha*K3*(q2^2+q1^2*K0) +
                       alpha*q1*q2*K4 + alpha*q1*q2*K5)/(q1+q2)^2/exp(probs_1[n][1]); //1->1
      int_1[n][2] = ((q2^2*lambda1+lambda0*q1*q2)*interevent[n] -
                       (lambda1*q1*q2+lambda0*q2^2)*K0*interevent[n] +
                       (lambda0-lambda1)*q2^2*K1 + (lambda1-lambda0)*q1*q2*K2 +
                       alpha*q2*K3*(q2-q1*K0) -
                       alpha*q2^2*K4 + alpha*q1*q2*K5)/(q1+q2)^2/exp(probs_1[n][2]); //2->1
      int_2[n][1] = ((q1*q2*lambda1+q1^2*lambda0)*interevent[n] -
                       (q1^2*lambda1+q1*q2*lambda0)*K0*interevent[n] +
                       (lambda1-lambda0)*q1^2*K1 + q1*q2*(lambda0-lambda1)*K2 +
                       alpha*q1*K3*(q2-q1*K0) +
                       alpha*q1^2*K4 - alpha*q2*q1*K5)/(q1+q2)^2/exp(probs_2[n][1]); //1->2
      int_2[n][2] = ((q1*q2*lambda1+lambda0*q1^2)*interevent[n] +
                       (q1*q2*lambda1+lambda0*q2^2)*K0*interevent[n] +
                       (lambda0-lambda1)*q1*q2*K1 + (lambda0-lambda1)*q1*q2*K2 +
                       alpha*q1*q2*K3*(1+K0) -
                       alpha*q1*q2*K4 - alpha*q1*q2*K5)/(q1+q2)^2/exp(probs_2[n][2]); //2->2
    }
    
    //consider n = 1
    forward[1][1] = log(lambda1) + log_sum_exp(probs_1[1]-int_1[1]+log_delta); 
    forward[1][2] = log(lambda0) + log_sum_exp(probs_2[1]-int_2[1]+log_delta); 
    
    if(num_events>1){
      for(n in 2:num_events){
        forward[n][1] = log_sum_exp(forward[n-1] + probs_1[n] - int_1[n]) + log(lambda1+alpha*R[n]);
        forward[n][2] = log_sum_exp(forward[n-1] + probs_2[n] - int_2[n]) + log(lambda0);
      }
    }
    
    forward_termination[1] = log_sum_exp(forward[num_events] + probs_1[num_events] - int_1[num_events]);
    forward_termination[2] = log_sum_exp(forward[num_events] + probs_2[num_events] - int_2[num_events]);
    // lots of places with max_Nm and Nm got rid of the +1
    target += log_sum_exp(forward_termination);
  }
}
"

```


# Fitting these models in RStan

To demonstrate how to use these models, we will simulate some data
from each of these models and fit the included `stan` models.


```{r sim-mmpp}
Q <- matrix(c(-0.4, 0.4, 0.2, -0.2), ncol = 2, byrow = TRUE)
mmpp_obj <- pp_mmpp(Q = Q, lambda0 = 1, c = 1.5, delta = c(1/3, 2/3))

sim_mmpp <- pp_simulate(mmpp_obj, n = 50)
```




We also show how to run these models using `rstan` for the MMPP 
example.

```{r rstan-mmpp}
mmpp_data <- list(num_events = length(sim_mmpp$events) - 1, 
                  # as first event is start time (0)
                  time_matrix = diff(c(sim_mmpp$events, sim_mmpp$end))
                  #interevent arrival time
                  )

mmpp_rstan <- stan(model_code = mmpp_stan_code, 
                   data = mmpp_data,
                   chains = 2)
mmpp_sim <- rstan::extract(mmpp_rstan)
```

The fit of this model can be evaluated in several ways, including
some of the tools in the `bayesplot` package, which are important to
ensure the fit is reasonable. 

```{r mmpp-plot}
bayesplot::mcmc_hist(as.matrix(mmpp_rstan), pars = c("lambda0", "c"))
```

To use the results of this fit with 
the functions of `ppdiag`, we want to extract an estimate
of each of the parameters. This can be achieved using the
posterior mean.

The desired posterior quantities can then be extracted from this object.

```{r mmpp-post}
mmpp_post <- lapply(mmpp_sim, mean)
```

Similarly, we can fit the MMHP model to simulated data also.

```{r sim-mmhp}
Q <- matrix(c(-0.4, 0.4, 0.2, -0.2), ncol = 2, byrow = TRUE)
mmhp_obj <- pp_mmhp(Q = Q, lambda0 = 0.5, lambda1 = 1.5,
                    alpha = 0.5, beta = 0.75, delta = c(1/3, 2/3))

sim_mmhp <- pp_simulate(mmhp_obj, n = 50)


mmhp_data <- list(num_events = length(sim_mmhp$events) - 1, 
                  # as first event is start time (0)
                  time_matrix = diff(c(sim_mmhp$events, sim_mmhp$end))
                  #interevent arrival time
                  )
```

```{r rstan-mmhp}
mmhp_rstan <- stan(model_code = mmhp_stan_code, 
                   data = mmhp_data,
                   chains = 2)
mmhp_sim <- rstan::extract(mmhp_rstan)
mmhp_post <- lapply(mmhp_sim, mean)
```

```{r plot-draws-mmhp}
bayesplot::mcmc_hist(as.matrix(mmhp_rstan),
                     pars = c("lambda0", "lambda1", "alpha", "beta"))
```



# Fitting these models in Cmdstanr

We can also fit these models using `cmdstanr`. We include the code
to fit these examples below (which is not run here).

```{r mmpp-fit, eval=FALSE}
mmpp_file <- write_stan_file(mmpp_stan_code)
mmpp_stan <- cmdstan_model(stan_file = mmpp_file)

fit_mmpp <- mmpp_stan$sample(data = mmpp_data,
                             seed = 123,
                             chains = 4,
                             parallel_chains = 4,
                             refresh = 500)
```

Similarly, we can fit the MMHP model in `cmdstanr`.

```{r compile-mmhp, eval=FALSE}
mmhp_file <- write_stan_file(mmhp_stan_code)
mmhp_stan <- cmdstan_model(stan_file = mmhp_file)

fit_mmhp <- mmhp_stan$sample(data = mmhp_data,
                             seed = 123,
                             chains = 4,
                             parallel_chains = 4,
                             refresh = 500)
```


# Using the Stan output with `ppdiag`

Having obtained estimates from fitting these stan models, we
can now pass these parameter estimates into `ppdiag` to 
evaluate model fit.

```{r mmpp-ppdiag, include = FALSE, eval=FALSE}
mmpp_fit_obj <- pp_mmpp(lambda0 = mmpp_post$lambda0,
                       c = mmpp_post$c,
                       Q = matrix( c(-mmpp_post$q1,
                                     mmpp_post$q1,
                                     mmpp_post$q2,
                                     -mmpp_post$q2),
                                   nrow = 2, ncol = 2, byrow = T) )


pp_diag(mmpp_fit_obj, events = sim_mmpp$events)

event_state <- ppdiag:::mmpp_event_state(params = mmpp_fit_obj,
                                         events = sim_mmpp$events[-1])
latent_inter <- ppdiag:::mmpp_latent(params = mmpp_fit_obj,
                                events = sim_mmpp$events[-1],
                                zt = event_state$zt)
# this looks like it's the problem here
# can write a test for this also, same for mmhp

pp_residual(object = mmpp_fit_obj, events = sim_mmpp$events)
```

```{r mmhp-ppdiag, fig.width=8}
mmhp_fit_obj <- pp_mmhp(lambda0 = mmhp_post$lambda0,
                        lambda1 = mmhp_post$lambda1,
                        alpha = mmhp_post$alpha,
                        beta = mmhp_post$beta,
                        Q = matrix( c(-mmhp_post$q1,
                                      mmhp_post$q1,
                                      mmhp_post$q2,
                                      -mmhp_post$q2),
                                 nrow = 2, ncol = 2, byrow = T) )

pp_diag(mmhp_fit_obj, events = sim_mmhp$events)

```
