---
title: "`ppdiag`, diagnostic tools for Point Processes"
output: rmarkdown::html_vignette
author: Sally Sun, Owen G. Ward et al.
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ppdiag)
```

This vignette provides an introduction to the functions 
available in `ppdiag` to evaluate the fit of univariate temporal 
point processes.

To achieve this, we currently include a range of functions which allow
a user to:

- Simulate data from a range of common univariate point processes.
- Fit a range of univariate point processes to data.
- After fitting a point process to some data, evaluate the ability of that 
point process to capture the temporal structure present in this data.


## Simulating data


## Fitting a point process


## Diagnosing the fit of a point process to data

There are several existing methods which can be used to assess the 
goodness of fit of a point process to temporal data. In this package 
we allow a user to:

- Examine the distribution of the rescaled interevent times, by utilising the
time rescaling theorem.
- Examine the residual process of an estimated point process, 
in particular computing the raw and Pearson residuals for a given point process fit to data.
- Visually inspect the estimated intensity of the point process.