---
title: 'ppdiag: a diagnostic tool for temporal data'
tags:
- R
- point processes
- event times
date: "`r Sys.Date()`"
output:
  rticles::joss_article: default
  pdf_document: 
    citation_package: biblatex
  html_document:
    df_print: paged
    citation_package: biblatex
authors:
- name: author names
  affiliation: 1
affiliations:
  name: Columbia University
  index: 1
bibliography: paper.bib
journal: JOSS
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(100) # to make it reproducible
```
# Summary
This package provides functions to evaluate the fit of univariate temporal point processes. 

Existing functions aim at: 

- Simulating data from a range of common univariate point processes including Homogeneous Poisson Process, Hawkes Process, and Markov-modulated Hawkes Process.

- Fitting a range of univariate point processes to data.

- After fitting a point process to some data, evaluating the ability of that 
point process to capture the temporal structure present in this data.

# Applying the functions on the Brastings dataset
```{r setup}
library(ppdiag)
library(SAPP)
data(Brastings)
```

## Fitting HPP to the Brastings data
```{r}
hpp = fithpp(Brastings)
```
## Plotting intensity of the fitted HPP
```{r}
drawHPPIntensity(hpp, Brastings, plot_events=TRUE)
```

## Diagnosing the fitted HPP
```{r}
diagpp(hpp, Brastings)
```

## Fitting HP to the Brastings data
```{r}
hp = fithp(events=Brastings)
```

## Plotting intensity of the fitted HP
```{r}
first_15=Brastings[1:15]
hp_15 = fithp(events=first_15)
drawHPIntensity(hp_15, events=first_15, plot_events=TRUE, int_title="HP Intensity for the first 15 Brastings events")
```

## Diagnosing the fitted HP
```{r}
diagpp(hp, Brastings)
```

In the network setting, Pearson residual is more appropriate in measuring the models.[@Jing2020] The diagnostics of the HPP model shows pearson residual=-13.58, and the diagnostics of the HP model shows pearson residual=14.5. It seems the HPP model is overestimating the intensity while the HP model is underestimating the intensity. We then look at the k-s statistics, the D statistic is 0.109 for HPP and 0.037 for HP, which indicates that HP is a better fit of the intensity since the D statistic is closer to 0. Additionally, the p-value for the HP model is much higher than the p-value for the HPP model, which also shows that the HP model is stronger in this case. This indicates that the Brastings data is better fitted to the Hawkes model. 

# Applying the functions on the email dataset
```{r, message=FALSE}
library(openintro)
library(lubridate)
first=ymd_hms(email$time[1],tz="EST")
second=ymd_hms(email$time[2],tz="EST")
time=(as.duration(interval(first, ymd_hms(email$time[-1],tz="EST")))/3600)[1:300]
```
## Fitting HPP to the email data
```{r}
hpp = fithpp(time)
```
## Plotting intensity of the fitted HPP
```{r}
drawHPPIntensity(hpp, time, plot_events=TRUE)
```
## Diagnosing the fitted HPP
```{r}
diagpp(hpp, time)
```
## Fitting HP to the email data
```{r}
#hp = fithp(vec=rep(0.5,3),events=time)
```
## Plotting intensity of the fitted HP
```{r}
#drawHPIntensity(hp, events=time, plot_events=TRUE)
```
## Diagnosing the fitted HP
```{r}
#diagpp(hp, time)
```

# References