---
title: "`ppdiag`, diagnostic tools for temporal Point Processes"
output: rmarkdown::html_vignette
author: Sally Sun, Owen G. Ward, Xiaoxi Zhao, Jing Wu, Tian Zheng.
vignette: >
  %\VignetteIndexEntry{`ppdiag`, diagnostic tools for temporal Point Processes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(100) # to make it reproducible
```

```{r setup}
# remotes::install_github("OwenWard/ppdiag")
library(ppdiag)
```

This vignette provides an introduction to the functions available in
`ppdiag` to evaluate the fit of univariate temporal point processes.

To achieve this, we currently include a range of functions which allow a
user to:

-   Simulate data from a range of common univariate point processes.
-   Fit a range of univariate point processes to data.
-   After fitting a point process to some data, evaluate the ability of
    that point process to capture the temporal structure present in this
    data.

## Classes

We create classes for each of the point process models included in the 
package. Currently, these are:

-   Homogeneous Poisson Process

`pp_hpp(lambda)` creates a `hpp` object with
rate parameter `lambda`.

```{r}
hpp_obj <- pp_hpp(lambda = 1)
hpp_obj
```

-   Hawkes Process:

`pp_hp(lambda0, alpha, beta, events = NULL)` creates a `hp` object.

```{r}
hp_obj <- pp_hp(lambda0 = 0.5, alpha = 0.2, beta = 0.5)
hp_obj
```

-   Markov Modulated Poisson Process:
`pp_mmpp(lambda0, lambda1, alpha, beta, Q, delta)` creates an `mmpp` object.

```{r}
Q <- matrix(c(-0.4, 0.4, 0.2, -0.2), ncol = 2, byrow = TRUE)

mmpp_obj <- pp_mmpp(Q, delta = c(1 / 3, 2 / 3), 
          lambda0 = 0.8,
          c = 1.2)

mmpp_obj
```

-   Markov-Modulated Hawkes Process:

`pp_mmhp(lambda0, lambda1, alpha, beta, Q, delta)` creates an `mmhp` object.

```{r}
mmhp_obj <- pp_mmhp(Q, delta = c(1 / 3, 2 / 3), 
          lambda0 = 0.2,
          lambda1 = .75,
          alpha = 0.1,
          beta = 0.2)

mmhp_obj

```

## Simulating data

To simulate data from a given point process, we use the function
`pp_simulate(pp_obj, ...)`. Here the first argument specifies one of the
above point processes, while the remaining arguments specify either the
number of events simulated or the length of the observation period for
possible events.

For example, we can simulate events up to a specified end time.

```{r}
hpp_events <- pp_simulate(hpp_obj, end = 10)
hpp_events
```

Alternatively, we can specify the number of events we wish to simulate.

```{r}
hp_events <- pp_simulate(hp_obj, start = 0, n = 20)
hp_events

```

This returns the simulated events of the specified point process. For
Markov Modulated processes, the states (and the times of these states)
are also returned. In this scenario only a specified number of events
can be simulated (currently).

```{r}
mmhp_events <- pp_simulate(object = mmhp_obj, n = 20)
mmhp_events
```

## Fitting a point process

For completeness, we include functions for fitting both homogeneous
Poisson and Hawkes processes to data. Fitting a Markov modulated model
is more complex, although we describe this procedure in an included
vignette.

`fithpp(hpp_events)` returns an object of class `hpp`, estimating the
MLE of a homogenous Poisson process for `hpp_events`

```{r}
fit_hpp <- fithpp(hpp_events)
fit_hpp
```

Similarly, `fithp(hp_events)` returns an object of class `hp`,
estimating the three parameters of the Hawkes process from `hp_events`
using `constrOptim`. This ensures that the returned solution (if one can
be obtained), satisfies the stationary condition of a Hawkes process.

```{r}
hp_events <- pp_simulate(hp_obj, n = 500)
fit_hp <- fithp(hp_events)
fit_hp$lambda0
fit_hp$alpha
fit_hp$beta
```

## Diagnosing the fit of a point process to data

The main goal of this package is to provide users with tools
to examine the fit of a specified point process to some data.
There are several methods which can be used to assess the
goodness of fit of a point process to temporal data. In this package we
allow a user to:

-   Visually inspect the estimated intensity of the point process.
-   Examine the fitted intensity along with the distribution of rescaled 
    inter-event times to help identify causes for lack of fit.
-   Examine the distribution of the rescaled inter-event times, by
    utilising the time rescaling theorem.
-   Examine the residual process of an estimated point process, in
    particular computing the raw and Pearson residuals for a given point
    process fit to data.


### Visualize the intensity function

`drawHPPIntensity(hpp, events)`
plots the intensity of a homogeneous Poisson process.

```{r,fig.width=4,fig.height=4}
drawHPPIntensity(fit_hpp, events = hpp_events,
                 color = "red")
```

Similarly, `drawHPIntensity(hp, events)`
plots the intensity of a Hawkes process. 

```{r,fig.width=4,fig.height=5}
drawHPIntensity(fit_hp, events = hp_events)
```

To plot the fitted intensity on the input events, set `fit=TRUE`.

```{r,fig.width=4,fig.height=5}
drawHPIntensity(events = hp_events, fit = TRUE)
```


Similarly, 
`drawUniMMHPIntensity(mmhp, mmhp_events)`
plots the intensity of a Markov modulated 
Hawkes process, with a similar 
function for Markov modulated Poisson processes. This requires both the point
process object and the output from `pp_simulate` which describes
the latent process.

```{r,fig.width=4,fig.height=5}
drawUniMMHPIntensity(mmhp_obj, mmhp_events)
```

### Visualize intensity and goodness of fit jointly

<!-- The main goal of this package is to provide tools to  -->
<!-- diagnose the fit of a given point process to data.  -->
<!-- To do this, we include several functions: -->

- `intensityqqplot` displays the estimated intensity of a given
point process along with a QQ-plot of the rescaled inter-event times. 
These together can often be useful in identifying issues with model fit for a chosen point process.

```{r intensityqqplot, fig.width=6, fig.height=5}
intensityqqplot(object = fit_hp, events = hp_events )
```



```{r intqqpot mmhp, eval=FALSE}
# this gives an error currently
intensityqqplot(object = mmhp_obj, events = mmhp_events$events)
```


### Residual Analysis

- `pp_residual` returns both raw and Pearson residuals from fitting
the specified point process to the given events.

```{r mmhp_residual}
pp_residual(object = mmhp_obj, events = mmhp_events$events)

pp_residual(object = fit_hp, events = hp_events)

```


### Overall summary of fit

- Finally, `pp_diag` summarises (both graphically and numerically) the fit
of a specified point process to the data. For a given point process
it computes the residuals (both raw and Pearson) obtained from fitting
that point process to the data, performs a goodness of fit test
based on the rescaled inter-event times, and displays graphical
summaries of this diagnostic.

```{r ppdiag hp, fig.width = 6, fig.height=4}
pp_diag(object = fit_hp, events = hp_events)
```


<!-- -   Homogeneous Poisson Process -->

<!-- `pp_diag(object, events)` gives diagnostics of the model, including a qq -->
<!-- plot, a ks plot and the corresponding ks test, -->
<!-- along with raw and Pearson residuals in one function. -->

<!-- ```{r,fig.width=7,fig.height=4} -->
<!-- pp_diag(hpp_obj,hpp_events) -->
<!-- ``` -->

<!-- `pp_residual(object, events)` -->
<!-- returns only the raw and Pearson residuals. -->

<!-- ```{r} -->
<!-- pp_residual(hpp_obj,hpp_events) -->
<!-- ``` -->

<!-- `intensityqqplot(object, events)` gives both qqplot and intensity plot. -->

<!-- ```{r,fig.width=7,fig.height=4} -->
<!-- intensityqqplot(hpp_obj, hpp_events) -->
<!-- ``` -->

<!-- -   Hawkes Process -->

<!-- `pp_diag(object, events)` gives diagnostics of the model, including a qq -->
<!-- plot, a ks plot, ks test, raw and pearson residuals in one function. -->

<!-- ```{r,fig.width=7,fig.height=4} -->
<!-- pp_diag(hp_obj,hp_events) -->
<!-- ``` -->

<!-- `pp_residual(object, events)` gives raw and pearson residuals. -->

<!-- ```{r} -->
<!-- pp_residual(hp_obj,hp_events) -->
<!-- ``` -->

<!-- `intensityqqplot(object, events)` gives both qqplot and intensity plots. -->

<!-- ```{r,fig.width=7,fig.height=4} -->
<!-- intensityqqplot(hp_obj, hp_events) -->
<!-- ``` -->

<!-- ### MMHP -->

<!-- ```{r,fig.width=7,fig.height=4} -->
<!-- pp_diag(mmhp_obj,mmhp_events$events) -->
<!-- ``` -->
