---
title: "PPDiag"
output: pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
devtools::load_all()  ## loads ppdiag
```

# Homogeneous Poisson Process

## Class
hpp(lambda, start=0, end=1, n = NULL) takes lambda(rate of the poisson process), start and end(start and end time of events), n(number of events) as inputs and returns an object of class hpp, in the form of a list of parameters provided. 
```{r}
pois_y <- hpp(lambda = 1, end = 10)

pois_y
```


## Simulate
simulatehpp takes a hpp object in the form of a list(lambda, start=0, end=1, n = NULL) and simulates events according to the arguments in the list, this function returns a vector of event times. 
```{r}
# events <- simulatehpp(lambda = 1, end = 10)
# events

pois_y$n <- 50
events <- simulatehpp(pois_y)
events


```


Below are several examples simulations(10 events, 50 events, and 100 events), their histogram of interevent times and the corresponding fitted density curves are plotted to illustrate homogeneous poisson events. 
```{r}
# sample simulation of 10 hpp events
s1=simulatehpp(hpp(lambda=1,n=10))
inter_s1=diff(s1)
hist(inter_s1,main="interevent times")
lines(density(inter_s1))


### can also just use diff(events) instead

# sample simulation of 50 hpp events
s2=simulatehpp(hpp(lambda=3,n=50))
inter_s2=diff(s2)
hist(inter_s2,main="interevent times")
lines(density(inter_s2))

# sample simulation of 100 hpp events
s3=simulatehpp(hpp(lambda=1,n=100))
inter_s3=diff(s3)
hist(inter_s3,main="interevent times")
lines(density(inter_s3))
```

## Fit
fithpp takes a vector of events and returns a hpp object in the form of list(lambda, start, end, n), events are fitted according to the maximum likelihood estimator.
```{r}

fithpp(events)

fit_hpp <- fithpp(events)

```

## Plot
plothpp_step takes in a vector of event times, a color parameter, and plots the event time counts.
```{r}
plothpp_step(events)
```

drawHPPIntensity takes (hpp, events = NULL, color = "red", 
                             plot_events = FALSE,
                             fit = FALSE,
                             int_title = "Intensity homogeneous Poisson Process"), which include a necessary argument which is a hpp object, and plots its intensity. You can also input 'events', which is a vector of event times, and indicate whether to plot the given events in 'plot_events', and indicate whether to fit a hpp on the given events or use the given hpp object in 'fit'.
```{r}
drawHPPIntensity(pois_y,color = "red")

# this plots the fit_hpp intensity and its corresponding simulated events, since plot_events default is FALSE:
drawHPPIntensity(fit_hpp,events=events, color = "red")

# this plots the inputted events, since plot_events=TRUE:
drawHPPIntensity(fit_hpp, events=events, color = "red", plot_events=TRUE)
```

## Diagnostics
diagpp takes a hpp object and a vector of events, and provides diagnostics of the model, including a qq plot, a ks plot, ks test, raw and pearson residuals. 
```{r}
diagpp(pois_y,t=events)
```

#This gives the right warning message, but we don't want to simulate data for
#these plots.

Intensityqqplot takes a hpp object, and a vector of events, and plots the qqplot and intensity plot together. 
```{r}
intensityqqplot(fit_hpp,fit_hpp$start,fit_hpp$end,
                history = 0,events = events)
```

## interevent times
Below is a plot of interevent times of our simulated events. 
```{r}
inter_hpp=diff(events)
hist(inter_hpp)
lines(density(inter_hpp))
```

# Hawkes Process

## Class
hp(lambda0, alpha, beta, events = NULL) takes lambda0(base intensity of hawkes process), alpha(jump size of increase in intensity), beta(exponential decay of intensity), events(the first event is at time zero, events data can be added later after simulation) as inputs and returns an object of class hp, in the form of a list of parameters provided.
```{r}
hp_obj <- hp(lambda0 = 0.1,alpha = 0.45,beta = 0.5)
```

## Simulate
```{r}
# simulatehp(lambda0 = 0.1, alpha = 0.5, beta = 0.5, start = 0, end = 10,
#           history = 0)

sims <- simulatehp(hp_obj,start = 0, end = 100, history = 0)
sims

```

## Fit
```{r}
init=rep(0.1,3)

fithp(init,sims$t,max(sims$t))

fit_hp <- fithp(init,sims$t,max(sims$t))
```

## Plot

One difficulty with this is that now we cannot look at the fits for different
parameter values....

```{r}
# changing this to use hawkes event times

drawHPIntensity(hp_obj, start=0, end=max(sims$t), history=0, events = sims$t,
                plot_events = TRUE)

drawHPIntensity(fit_hp, start=0, end=max(sims$t), 
                history=0, events = sims$t, plot_events = TRUE)

```

## Diagnostics
```{r}
diagpp(hp_obj, t=sims$t)
# this is the truth


diagpp(fit_hp, t = sims$t)

```


```{r}
intensityqqplot(fit_hp,start=0,end=max(sims$t),history=0,events=events)
```

## interevent times
```{r}
temp=sims$t
inter_hp=sims$t[2:length(sims$t)]-temp[1:length(temp)-1]
hist(inter_hp)
```

# MMHP
## Class
mmhp(lambda0, lambda1, alpha, beta, Q = NULL, delta = NULL, events = NULL) takes lambda0(intensity for hpp), lambda1(base intensity for hp), alpha(jump size of the increase in intensity in hp), beta(exponential decrease of intensity in hp), Q(transition probability matrix), delta(initial state probability), events(event times, could be added later after simulation) as inputs and returns an object of class mmhp, in the form of a list of parameters provided.
```{r}
Q <- matrix(c(-0.4, 0.4, 0.2, -0.2), ncol = 2, byrow = TRUE)

x <- mmhp(Q, delta = c(1 / 3, 2 / 3), 
          lambda0 = 0.2,
          lambda1 = .75,
          alpha = 0.1,
          beta = 0.2)
```

## Simulate
```{r}
y <- simulatemmhp(x,n = 50)
```

## Plot
```{r}
drawUniMMHPIntensity(x, y)
```

## Diagnostics
```{r}
try(diagpp(x, t=y))
try(intensityqqplot(object=x,min(y$events),max(y$events),history=0,events=y$events,pzt=0.5))
```
